\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-11-10 metatheory-condensed.tex
add one sentence about parser case study
edits in FMLTT sec

\TT is intended as a foundational model. So unlike our programmer-facing plugin,
typing is structural rather than nominal, and
partial recursors and propositional equalities are not automated but have to
be explicitly defined.
\TT provides MLTT-style constructs that can be used to \emph{encode}
family polymorphism rather than building it in directly.
%
Most notably, \TT extends MLTT with what we call \emph{linkages}.\footnote{
Linkages are a namesake of the theoretical device through which prior work~\cite{zm2017} models
family polymorphism in an OO setting, but the technical details differ significantly.}
Linkages model families, so they are tuples composed of fields but with a twist: they
support late binding.
Unlike dependent tuples where a later component is \emph{existentially}
quantified over the earlier ones, a linkage component is \emph{universally}
quantified over---and thus polymorphic to---components preceding it.
%
\TT also features \emph{linkage transformers} that can be used to inductively
construct linkages from other linkages, by inheriting or overriding components of a linkage,
or by extending a linkage with new components.
%
Inductive types are modeled as W-types~\cite{martin1984intuitionistic} and their
extension as overriding.
%
Consistency and canonicity are proved by giving semantic models to the syntax.

@@

\noindentparagraph{MLTT.}

\cref{fig:mltt-selected} presents the syntax and selected typing rules of MLTT, and
\cref{fig:fmltt-selected} \TT.
%The new forms (relative to MLTT) are highlighted.
We briefly review the base MLTT fragment first.
%
As is standard in MLTT presentations, we use explicit substitutions~\cite{substcalculus,abadi1989subst}:
substitutions~$\gm$ and their applications (e.g., \sub{T}{\gm}) are part
of the syntax rather than meta-operations.
%
Variables are represented by de Bruijn indices:
\var{n} is the variable bound by the $n$-th closest enclosing binder.
For example, $\lambda x.\,\lambda y.\,x$ is $\lam{\lam{\var{1}}}$.
%
Typing judgments for substitutions have the form 
%\EDJ{It seems to me G |- g : D is a judgement not a rule. Rules usually have two parts (premises + conclusion). Judgment is a standard terminology invented by Martin Lof (https://ncatlab.org/nlab/show/judgment see reference here). I will suggest saying 'Formation rules for substitution lead to a judgement G |- g : D' sth like this. In pen-and-paper, the formation rule (corresponding to QIIT well-typed-ness of judgement) is called presuppose (can be found in Modal MLTT paper) }
$\goodSub{\Gm}{\gm}{\Dl}$.
The idea is that applying~$\gm$ to terms valid in the context~$\Dl$
yields terms valid in~$\Gm$ (e.g., rule \ruleref{tm/sub}).
%
The two main forms of substitutions are weakening (\ruleref{sub/wk}) and extension (\ruleref{sub/ext}):
\sub{t}{\SubstWeak{n}} introduces $n$ free variables into the context of $t$, and
\sub{t}{\SubstExt{\gm}{t'}} substitutes~$t'$ for~$\var{0}$ in~$t$ and
then applies~$\gm$.
For example, rule \ruleref{tm/snd} states that if $t$ is a dependent pair
that has type $\TySigma{A}{B}$, then $\snd{t}$ has type
$\sub{B}{\SubstExt{\SubstWeak{0}}{\fst{t}}}$.

@@

Rules \ruleref{tm/wsup} and \ruleref{tm/wrec} require knowing~$\wsig$:
the W-type is exhaustively generated by its constructors, and
its elimination must exhaustively handle all the constructors in
its signature.

@@

This type~$A$ abstracting the context in \ruleref{l/add} and
\ruleref{lsig/add} is not necessarily the same as those recorded by~$\lsig$,
because a field representing the code $\wcode{\wsig}$ of a W-type has to
expose different types to different succeeding fields.
Succeeding fields that invoke
$\Wsup{\wsig}{\cDot}{\cDot}$ or $\Wrec{\wsig}{\cDot}{\cDot}$
should see the concrete signature~$\wsig$, as rules \ruleref{tm/wsup}
and \ruleref{tm/wrec} stipulate, whereas $\wsig$ should be hidden
from all other fields so that they can be reused
for a W-type signature~$\wsig'$ extending~$\wsig$.

@@

\item
Similarly, $\wsig_{\texttt{tm}}$ is exposed to the typing context of
$t_{10}$, which models the recursive function \lsti{subst} by invoking
the recursor $\Wrec{\wsig_{\texttt{tm}}}{\cDot}{\cDot}$.
%
Partial recursors automatically generated by the plugin
can also be modeled in this way by invoking $\texttt{Wrec}$.

>>>>

\TT is intended as a foundational model. So unlike our programmer-facing plugin,
typing is structural rather than nominal, and
fields automatically axiomatized by the plugin require explicit definitions in \TT.
%
\TT provides MLTT-style constructs that can be used to express families and
family polymorphism. %rather than building it in directly.
%
Most notably, it extends MLTT with what we call \emph{linkages}.
(Linkages are a namesake of the theoretical device through which prior work~\cite{zm2017} models
family polymorphism in an OO setting, but the technical details differ significantly from the prior work.)

\begin{itemize}
[labelsep=*,leftmargin=1pc,itemsep=3pt]
\item 
Linkages model families, so they are like tuples composed of the fields.
But there is a twist: they support late binding.
Unlike dependent tuples where a later component is \emph{existentially}
quantified over the earlier ones, a linkage component is \emph{universally}
quantified over---and thus polymorphic to---components preceding it.

\item
\TT also features \emph{linkage transformers} that enable inductive
construction of linkages (and thus families): a linkage can be constructed by inheriting or
overriding components of an existing linkage, or by extending an existing
linkage with new components.

\item
Inductive types are modeled as W-types~\cite{martin1984intuitionistic} and their
extension as overriding.

\item
Consistency and canonicity of \TT are proved by giving semantic interpretations to the syntactic typing judgments.
\end{itemize}

@@

\noindentparagraph{Brief review of MLTT.}

\cref{fig:mltt-selected} presents the syntax and selected typing rules of MLTT
(and \cref{fig:fmltt-selected} \TT).
%We briefly review the base MLTT fragment first.
%
As is conventional with MLTT presentations, we use \emph{explicit substitutions}~\cite{substcalculus,abadi1989subst}:
substitutions~$\gm$ and their applications (e.g., \sub{T}{\gm}) are part
of the syntax rather than meta-operations.
%
Variables are represented by de Bruijn indices:
\var{n} is the variable bound by the $n$-th closest enclosing binder.
For example, $\lambda x.\,\lambda y.\,x$ is $\lam{\lam{\var{1}}}$.
%
Typing judgments for substitutions have the form 
%\EDJ{It seems to me G |- g : D is a judgement not a rule. Rules usually have two parts (premises + conclusion). Judgment is a standard terminology invented by Martin Lof (https://ncatlab.org/nlab/show/judgment see reference here). I will suggest saying 'Formation rules for substitution lead to a judgement G |- g : D' sth like this. In pen-and-paper, the formation rule (corresponding to QIIT well-typed-ness of judgement) is called presuppose (can be found in Modal MLTT paper) }
$\goodSub{\Gm}{\gm}{\Dl}$.
The idea is that applying~$\gm$ to terms valid in the context~$\Dl$
yields terms valid in~$\Gm$ (e.g., rule \ruleref{tm/sub}).
%
The two main forms of substitutions are weakening (\ruleref{sub/wk}) and extension (\ruleref{sub/ext}):
\sub{t}{\SubstWeak{n}} introduces $n$ free variables into the context of $t$, and
\sub{t}{\SubstExt{\gm}{t'}} substitutes~$t'$ for~$\var{0}$ in~$t$ and
then applies~$\gm$.
For example, rule \ruleref{tm/snd} states that if $t$ is a dependent pair
that has type $\TySigma{A}{B}$, then $\snd{t}$ has type
$\sub{B}{\SubstExt{\SubstWeak{0}}{\fst{t}}}$,
where $\SubstId$ is the identity substitution (\ruleref{sub/id}).

@@

Rules \ruleref{tm/wsup} and \ruleref{tm/wrec} require access to~$\wsig$:
the W-type is exhaustively generated by its constructors, and
its elimination must exhaustively handle all the constructors in
its signature.

@@

This type~$A$ abstracting the context in \ruleref{l/add} and
\ruleref{lsig/add} is not necessarily the same as those recorded by~$\lsig$,
because a field representing the code $\wcode{\wsig}$ of a W-type has to
expose different types to different fields that come after it.
Later fields that invoke
$\Wsup{\wsig}{\cDot}{\cDot}$ or $\Wrec{\wsig}{\cDot}{\cDot}$
should see the concrete signature~$\wsig$, as rules \ruleref{tm/wsup}
and \ruleref{tm/wrec} stipulate, whereas $\wsig$ should be hidden
from all other fields so that they can be reused
for a W-type signature~$\wsig'$ extending~$\wsig$.

@@

\item
Similarly, $\wsig_{\texttt{tm}}$ is exposed to the typing context of
$t_{10}$, which models the recursive function \lsti{subst} by invoking
the recursor $\Wrec{\wsig_{\texttt{tm}}}{\cDot}{\cDot}$.
%
Like \lsti{subst}, partial recursors in \Lang are axiomatized by the plugin
(\cref{sec:latebind,sec:prec}), and they can similarly be defined in \TT
using $\texttt{Wrec}$.
